---
title: "Assignment #3"
author: "Puneet Velidi"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["amsmath","amssymb","amsthm"]
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(deSolve)
```

\newpage

# Q1 – Boundary Value Problem (10 points)

We're given the BVP

$$
(a(x)\, y')' = f(x), \quad 0 \le x \le 1, \quad y(0) = 1,\; y(1) = 2,
$$

with $a(x) = 1 + x^2$ and $f(x) = 2 + 6x^2$.

---

## (a) Uniqueness and exact solution

### Uniqueness (energy method)

Assume two solutions $y_1, y_2$ exist and see what happens. Let $w = y_1 - y_2$. Then $w$ satisfies:

$$
(a(x)\,w')' = 0, \quad w(0) = 0,\; w(1) = 0.
$$

Multiply both sides by $w$ and integrate:

$$
\int_0^1 w\,(a(x)\,w')'\,dx = 0.
$$

Integrate by parts. The boundary term $[w\,a(x)\,w']_0^1$ vanishes since $w(0) = w(1) = 0$, so we're left with:

$$
-\int_0^1 a(x)\,(w')^2\,dx = 0
\quad \Longrightarrow \quad
\int_0^1 a(x)\,(w')^2\,dx = 0.
$$

Now, $a(x) = 1 + x^2 > 0$ on $[0,1]$ and $(w')^2 \ge 0$, so the only way this integral is zero is if $w' = 0$ everywhere:

$$
w'(x) = 0 \;\;\forall\, x \;\;\Longrightarrow\;\; w \equiv \text{const} \;\;\Longrightarrow\;\; w \equiv 0 \quad (\text{since } w(0)=0).
$$

So $y_1 = y_2$. By the existence theorem for linear BVPs with smooth, positive $a(x)$, a solution exists — together with uniqueness, we get exactly one solution. $\blacksquare$

### Checking that $y_e(x) = x^2 + 1$ is the solution

Just plug it in. We have $y_e' = 2x$, so:

$$
a(x)\,y_e' = (1+x^2)(2x) = 2x + 2x^3
$$

$$
\frac{d}{dx}\big[a(x)\,y_e'\big] = 2 + 6x^2 = f(x) \quad \checkmark
$$

Boundary conditions: $y_e(0) = 0 + 1 = 1$ $\checkmark$, $\;y_e(1) = 1 + 1 = 2$ $\checkmark$.

Both check out. So $y_e(x) = x^2 + 1$ is the unique solution. $\blacksquare$

---

## (b) Discretization using symmetric differencing

Set up a grid with $n$ interior points: $x_i = ih$, $h = 1/(n+1)$, for $i = 0, \dots, n+1$. Define the half-point coefficients $a_{i\pm1/2} = a(x_i \pm h/2)$.

The idea is to approximate $\frac{d}{dx}[a(x)y']$ in two stages. First, centred differences at the half-points for the outer derivative:

$$
\frac{d}{dx}[a(x)y']\bigg|_{x_i}
\approx \frac{1}{h}\Big[a_{i+1/2}\,y'(x_i+h/2) - a_{i-1/2}\,y'(x_i-h/2)\Big]
$$

Then approximate each $y'$ at the half-points:

$$
y'(x_i + h/2) \approx \frac{y_{i+1} - y_i}{h}, \qquad
y'(x_i - h/2) \approx \frac{y_i - y_{i-1}}{h}
$$

Putting it together, at each interior node $i = 1, \dots, n$:

$$
\frac{1}{h^2}\Big[a_{i-1/2}\,y_{i-1}
- (a_{i+1/2} + a_{i-1/2})\,y_i
+ a_{i+1/2}\,y_{i+1}\Big] = f(x_i)
$$

This gives us a linear system $A_h Y = F$ with $Y = (y_1,\dots,y_n)^T$. The matrix $A_h$ is tridiagonal:

$$
(A_h)_{i,i} = -\frac{a_{i+1/2}+a_{i-1/2}}{h^2}, \quad
(A_h)_{i,i-1} = \frac{a_{i-1/2}}{h^2}, \quad
(A_h)_{i,i+1} = \frac{a_{i+1/2}}{h^2}
$$

The boundary values $y(0) = 1$ and $y(1) = 2$ get absorbed into the first and last equations:

$$
F_1 = f(x_1) - \frac{a_{1/2}}{h^2}\,y(0), \qquad
F_n = f(x_n) - \frac{a_{n+1/2}}{h^2}\,y(1), \qquad
F_i = f(x_i) \;\;(2 \le i \le n-1)
$$

---

## (c) Showing the truncation error is $O(h^2)$

We need to show that plugging the exact solution $y_e$ into our difference formula leaves a residual that goes to zero like $h^2$. The truncation error at node $x_i$:

$$
\tau_i = \frac{1}{h^2}\Big[a_{i-1/2}\,y_e(x_{i-1})
- (a_{i+1/2}+a_{i-1/2})\,y_e(x_i)
+ a_{i+1/2}\,y_e(x_{i+1})\Big] - f(x_i)
$$

Expand everything in Taylor series around $x_i$:

$$
y_e(x_i \pm h) = y_e \pm hy_e' + \frac{h^2}{2}y_e'' \pm \frac{h^3}{6}y_e''' + \frac{h^4}{24}y_e^{(4)} + \cdots
$$

$$
a_{i\pm 1/2} = a(x_i) \pm \frac{h}{2}a'(x_i) + \frac{h^2}{8}a''(x_i) \pm \cdots
$$

Multiply these out, substitute, and collect by powers of $h$:

- The $O(h^{-2})$ and $O(h^{-1})$ terms cancel out
- The $O(1)$ terms give $(a\,y_e')' = f(x_i)$, which cancels with $-f(x_i)$
- What's left starts at $O(h^2)$

$$
\therefore \quad \tau_i = O(h^2), \qquad \|\tau_h\|_\infty = O(h^2) \qquad \blacksquare
$$

---

## (d) Showing $A_h$ is non-singular

We want to show the system always has a unique solution. The approach is to show $-A_h$ is positive definite.

Take any nonzero $X = (x_1,\dots,x_n)^T$ and set $x_0 = x_{n+1} = 0$ (homogeneous boundary conditions). Doing a discrete summation by parts on $-X^T A_h X$:

$$
-X^T A_h X = \frac{1}{h^2}\sum_{i=0}^{n} a_{i+1/2}\,(x_{i+1} - x_i)^2
$$

Every term is non-negative since $a_{i+1/2} = 1 + (x_i + h/2)^2 > 0$. The sum can only be zero if $x_{i+1} = x_i$ for all $i$, but since $x_0 = 0$ that would force $X = 0$:

$$
-X^T A_h X > 0 \;\;\forall\, X \ne 0
\quad \Longrightarrow \quad -A_h \text{ is positive definite}
\quad \Longrightarrow \quad A_h \text{ is non-singular} \qquad \blacksquare
$$

---

## (e) Numerical experiments

Now let's solve the system for $h = 1/32, 1/64, 1/128$ and check that the error behaves like $O(h^2)$.

```{r q1e-solve}
exact_sol <- function(x) x^2 + 1
a_fun     <- function(x) 1 + x^2
f_fun     <- function(x) 2 + 6 * x^2

solve_bvp <- function(n) {
  h  <- 1 / (n + 1)
  xi <- (1:n) * h

  ap <- a_fun(xi + h / 2)
  am <- a_fun(xi - h / 2)

  d_main  <- -(ap + am) / h^2
  d_upper <- ap[1:(n - 1)] / h^2
  d_lower <- am[2:n]       / h^2

  A <- diag(d_main)
  for (k in 1:(n - 1)) {
    A[k, k + 1] <- d_upper[k]
    A[k + 1, k] <- d_lower[k]
  }

  rhs    <- f_fun(xi)
  rhs[1] <- rhs[1] - am[1] / h^2 * 1
  rhs[n] <- rhs[n] - ap[n] / h^2 * 2

  Y <- solve(A, rhs)
  list(x = xi, Y = Y, h = h)
}

h_vals <- c(1/32, 1/64, 1/128)
n_vals <- round(1 / h_vals) - 1
errors <- numeric(length(h_vals))

for (k in seq_along(n_vals)) {
  sol       <- solve_bvp(n_vals[k])
  errors[k] <- max(abs(sol$Y - exact_sol(sol$x)))
}

rates <- log(errors[1:(length(errors) - 1)] / errors[2:length(errors)]) / log(2)

results <- data.frame(
  h   = h_vals,
  n   = n_vals,
  e_h = errors,
  r_h = c(NA, rates)
)

knitr::kable(results, digits = 10,
             col.names = c("h", "n", "e_h", "r_h"),
             caption = "Errors and convergence rates")
```

We get $r_h \approx 2$, confirming the scheme is second-order accurate. Each time we halve $h$, the error drops by roughly a factor of 4 — exactly what we'd expect from $e_h = O(h^2)$.

\newpage

# Q2 – Eigenvalues of a constant-coefficient tridiagonal matrix (5 points)

We have the $n \times n$ tridiagonal matrix with constant $a$ on the diagonal, $c > 0$ on the sub-diagonal, and $b > 0$ on the super-diagonal:

$$
A = \begin{pmatrix}
a & b \\
c & a & b \\
  & c & \ddots & \ddots \\
  &   & \ddots & a & b \\
  &   &        & c & a
\end{pmatrix}
$$

## Eigenvalues via difference-scheme method

Writing out $AV = \lambda V$ for the $j$-th row:

$$
c\,v_{j-1} + a\,v_j + b\,v_{j+1} = \lambda\,v_j, \qquad j = 1, \dots, n
$$

We treat this as a recurrence with boundary conditions $v_0 = 0$, $v_{n+1} = 0$. Try the ansatz:

$$
v_j = \left(\frac{b}{c}\right)^{j/2} \sin(j\theta)
$$

This gives $v_0 = \sin(0) = 0$, which is good. Plugging in and using the identity $\sin((j-1)\theta) + \sin((j+1)\theta) = 2\cos\theta\sin(j\theta)$:

$$
2\sqrt{bc}\cos\theta\,\sin(j\theta) + a\sin(j\theta) = \lambda\sin(j\theta)
$$

Since $\sin(j\theta) \ne 0$ for the eigenvectors we care about, we can divide through:

$$
\lambda = a + 2\sqrt{bc}\,\cos\theta
$$

For the other boundary condition $v_{n+1} = 0$, we need $\sin((n+1)\theta) = 0$, giving:

$$
\theta_k = \frac{k\pi}{n+1}, \qquad k = 1, 2, \dots, n
$$

### Results

**Eigenvalues:**

$$
\boxed{\lambda_k = a + 2\sqrt{bc}\,\cos\!\left(\frac{k\pi}{n+1}\right), \quad k = 1, \dots, n}
$$

**Eigenvectors:** The $k$-th eigenvector has components:

$$
v_j^{(k)} = \left(\frac{b}{c}\right)^{j/2} \sin\!\left(\frac{jk\pi}{n+1}\right), \quad j = 1, \dots, n
$$

**When is $A$ non-singular?** All eigenvalues must be nonzero:

$$
a \ne -2\sqrt{bc}\,\cos\!\left(\frac{k\pi}{n+1}\right) \quad \forall\, k = 1, \dots, n
$$

Since $|\cos(k\pi/(n+1))| < 1$, a sufficient condition is $|a| > 2\sqrt{bc}$. $\blacksquare$

\newpage

# Q3 – Shooting method for a nonlinear BVP (10 points)

We want to solve:

$$
v'' = x^2 + 3v + 10v^3, \quad 0 \le x \le 1, \quad v(0) = 0,\; v(1) = \beta
$$

### Setting up the IVP

Convert to a first-order system with $y_1 = v$, $y_2 = v'$:

$$
y_1' = y_2, \qquad y_2' = x^2 + 3y_1 + 10y_1^3
$$

The shooting idea: pick $v'(0) = \alpha$, solve the IVP, and adjust $\alpha$ until $v(1) = \beta$.

For Newton's method we also need the variational equation. Letting $w = \partial v / \partial \alpha$:

$$
w'' = (3 + 30v^2)\,w, \qquad w(0) = 0,\; w'(0) = 1
$$

The Newton update is:

$$
\alpha_{k+1} = \alpha_k - \frac{v(1;\alpha_k) - \beta}{w(1;\alpha_k)}
$$

```{r q3-setup}
ydotp_r <- function(t, y, parms) {
  dy1 <- y[2]
  dy2 <- 3 * y[1] + t^2 + 10 * y[1]^3
  dy3 <- y[4]
  dy4 <- (3 + 30 * y[1]^2) * y[3]
  list(c(dy1, dy2, dy3, dy4))
}

ydot_r <- function(t, y, parms) {
  dy1 <- y[2]
  dy2 <- 3 * y[1] + t^2 + 10 * y[1]^3
  list(c(dy1, dy2))
}
```

---

## (a) Newton shooting – finding $\beta_c$

We use Newton with initial guess $\alpha = -1$ and try different values of $\beta$. Something goes wrong between $\beta = 3.5$ and $\beta = 4$, so we bisect on $\beta$ to pin down where Newton stops converging.

```{r q3a-newton}
newton_shooting <- function(beta, alpha0 = -1, tol = 1e-10, maxiter = 100) {
  s <- alpha0
  converged <- FALSE
  for (i in 1:maxiter) {
    sol <- tryCatch(
      ode(y = c(0, s, 0, 1), times = seq(0, 1, length.out = 500),
          func = ydotp_r, parms = NULL, method = "ode45"),
      error = function(e) NULL
    )
    if (is.null(sol) || any(!is.finite(sol[, 2]))) return(list(converged = FALSE))
    v1 <- tail(sol[, 2], 1)
    w1 <- tail(sol[, 4], 1)
    if (!is.finite(v1) || !is.finite(w1) || abs(w1) < 1e-15)
      return(list(converged = FALSE))
    s_new <- s - (v1 - beta) / w1
    if (!is.finite(s_new)) return(list(converged = FALSE))
    if (abs(s_new - s) < tol) { s <- s_new; converged <- TRUE; break }
    s <- s_new
  }
  if (!converged) return(list(converged = FALSE))
  sol_f <- ode(y = c(0, s), times = seq(0, 1, length.out = 500),
               func = ydot_r, parms = NULL, method = "ode45")
  list(x = sol_f[, 1], v = sol_f[, 2], alpha = s, converged = TRUE)
}

test_newton <- function(beta) {
  res <- newton_shooting(beta)
  res$converged && all(is.finite(res$v))
}

beta_lo <- 3.5; beta_hi <- 4.0
for (iter in 1:50) {
  beta_mid <- (beta_lo + beta_hi) / 2
  if (test_newton(beta_mid)) beta_lo <- beta_mid else beta_hi <- beta_mid
}
beta_c <- (beta_lo + beta_hi) / 2
cat(sprintf("Estimated beta_c = %.4f\n", beta_c))
```

```{r q3a-plot, fig.width=7, fig.height=5}
betas_plot <- c(3.0, 3.3, round(beta_c - 0.02, 2))
plot(NULL, xlim = c(0, 1), ylim = c(0, 5),
     xlab = "x", ylab = "v(x)",
     main = "Newton shooting: solutions for beta < beta_c")
cols <- c("blue", "red", "darkgreen")
for (k in seq_along(betas_plot)) {
  res <- newton_shooting(betas_plot[k])
  if (res$converged) lines(res$x, res$v, col = cols[k], lwd = 2)
}
legend("topleft", legend = paste("beta =", betas_plot), col = cols, lwd = 2, cex = 0.9)
```

So Newton works fine for $\beta < \beta_c \approx `r sprintf("%.3f", beta_c)`$, but blows up past that. You might think the BVP just doesn't have a solution for $\beta > \beta_c$ — but as we'll see in part (b), that's not quite right.

---

## (b) Bisection shooting and what's really going on

Now let's try bisection instead (using R's `uniroot`). Rather than using derivatives, bisection just brackets the root and narrows in.

```{r q3b-bisection}
shoot_residual <- function(alpha, beta) {
  sol <- tryCatch(
    ode(y = c(0, alpha), times = seq(0, 1, length.out = 2000),
        func = ydot_r, parms = NULL, method = "ode45"),
    error = function(e) NULL
  )
  if (is.null(sol) || any(!is.finite(sol[, 2]))) return(1e10 * sign(alpha))
  tail(sol[, 2], 1) - beta
}

bisection_shooting <- function(beta, lo = -10, hi = 10) {
  tryCatch({
    alpha_star <- uniroot(shoot_residual, interval = c(lo, hi),
                          beta = beta, tol = 1e-10, maxiter = 2000)$root
    sol <- ode(y = c(0, alpha_star), times = seq(0, 1, length.out = 500),
               func = ydot_r, parms = NULL, method = "ode45")
    list(x = sol[, 1], v = sol[, 2], alpha = alpha_star, converged = TRUE)
  }, error = function(e) list(converged = FALSE))
}

betas_test <- c(3.5, 3.8, 4.0, 4.5)
cat("Bisection shooting results:\n")
for (b in betas_test) {
  res <- bisection_shooting(b)
  if (res$converged) {
    cat(sprintf("  beta = %.1f : converged, alpha = %.6f, v(1) = %.6f\n",
                b, res$alpha, tail(res$v, 1)))
  } else {
    cat(sprintf("  beta = %.1f : did not converge\n", b))
  }
}
```

Bisection finds solutions even for $\beta > \beta_c$. So the problem isn't that solutions don't exist; it's that Newton can't find them.

### Why Newton fails

The hint says to look at the IVP with $v'(0) = 1.236$ on $[0, 0.999]$:

```{r q3b-hint, fig.width=7, fig.height=5}
sol_hint <- ode(y = c(0, 1.236), times = seq(0, 0.999, length.out = 2000),
                func = ydot_r, parms = NULL, method = "ode45")
plot(sol_hint[, 1], sol_hint[, 2], type = "l", lwd = 2, col = "blue",
     xlab = "x", ylab = "v(x)",
     main = expression(paste("IVP with v(0)=0, v'(0)=1.236 on [0, 0.999]")))
abline(v = 1, lty = 2, col = "gray")
```

The solution shoots up steeply near $x = 1$. What's happening is that for $\alpha$ values producing large $\beta$, the solution develops a near-singularity:

$$
v(x;\alpha) \to \infty \text{ as } x \to 1
\quad \Longrightarrow \quad
w(1) = \frac{\partial v(1)}{\partial \alpha} \to \infty
$$

Newton relies on $w(1)$ to compute $\Delta\alpha = -(v(1)-\beta)/w(1)$. When $w(1)$ blows up, the Newton steps become unreliable and the iteration diverges. Bisection doesn't have this problem — it only needs the sign of $v(1;\alpha) - \beta$, not its derivative. That's why bisection is more robust here.

\newpage

# Q4 – Invertibility of a variable-coefficient tridiagonal matrix (5 points)

We have an $n \times n$ tridiagonal matrix $A$ with diagonal entries $a_i$, sub-diagonal $b_i$ (in row $i$), super-diagonal $c_i$ (in row $i$), satisfying:

$$
|b_i| + |c_i| \le |a_i| \;\;(i=2,\dots,n-1), \qquad |c_1| < |a_1|, \qquad |b_n| < |a_n|
$$

**Goal:** Show $A$ is invertible.

## Proof by contradiction

Suppose $A$ is not invertible, so there's some nonzero $X$ with $AX = 0$. Normalize so $\|X\|_\infty = 1$. We'll show this leads to a contradiction.

Writing out $AX = 0$ row by row:

$$
\text{Row 1:} \quad a_1 x_1 + c_1 x_2 = 0
$$
$$
\text{Row } i: \quad b_i x_{i-1} + a_i x_i + c_i x_{i+1} = 0 \qquad (2 \le i \le n-1)
$$
$$
\text{Row } n: \quad b_n x_{n-1} + a_n x_n = 0
$$

**Step 1 — show $|x_1| < 1$:** From Row 1, $a_1 x_1 = -c_1 x_2$, so:

$$
|a_1|\,|x_1| = |c_1|\,|x_2| \le |c_1| \cdot 1 < |a_1|
\quad \Longrightarrow \quad |x_1| < 1
$$

We used $|x_2| \le \|X\|_\infty = 1$ and the strict condition $|c_1| < |a_1|$.

**Step 2 — show $|x_i| < 1 \Rightarrow |x_{i+1}| < 1$:** For $1 \le i \le n-2$, from Row $i+1$:

$$
|a_{i+1}|\,|x_{i+1}| \le |b_{i+1}|\,|x_i| + |c_{i+1}|\,|x_{i+2}|
$$

We know $|x_i| < 1$ from the inductive hypothesis, and $|x_{i+2}| \le 1$. The key strict inequality:

$$
|b_{i+1}|\,|x_i| + |c_{i+1}| < |b_{i+1}| + |c_{i+1}| \le |a_{i+1}|
\quad \Longrightarrow \quad |x_{i+1}| < 1
$$

For $i = n-1$, from the last row:

$$
|a_n|\,|x_n| = |b_n|\,|x_{n-1}| \le |b_n| < |a_n|
\quad \Longrightarrow \quad |x_n| < 1
$$

**Putting it together:** By induction,

$$
|x_1| < 1 \;\Rightarrow\; |x_2| < 1 \;\Rightarrow\; \cdots \;\Rightarrow\; |x_n| < 1
\quad \Longrightarrow \quad \|X\|_\infty < 1
$$

But we assumed $\|X\|_\infty = 1$ — contradiction. Therefore $A$ is invertible. $\blacksquare$
