---
title: "RPCB"
output: html_document
date: "2025-12-12"
---

```{r}
library(dplyr)
library(readr)
library(ggplot2)
library(patchwork)
if (!require("ejabT1E", quietly = TRUE)) {
  stop("Install ejabT1E: devtools::install_github(\"nathoogroup/ejabT1E\", subdir = \"ejabT1E\")")
}
# diagnostic_qqplot exists in ejabT1E but is not exported
diagnostic_qqplot <- getFromNamespace("diagnostic_qqplot", "ejabT1E")
```

```{r}
alpha = 0.05
```

```{r setup, include=FALSE}
rpcb = read.csv("RP_CB Final Analysis - Effect level data.csv")
rpcb_df = rpcb[, c("Paper..", "Original.sample.size", "Original.p.value",
                    "Replication.p.value", "Original.test.statistic.type",
                    "Original.df1", "Original.df2")]
rpcb_df = rpcb_df[!is.na(rpcb_df$Original.p.value) &
                   !is.na(rpcb_df$Original.sample.size) &
                   !is.na(rpcb_df$Replication.p.value), ]
rpcb_df$replicated = ifelse(rpcb_df$Replication.p.value < alpha, 1, 0)
```

Core eJAB functions (from ejabT1E: ejab01, detect_type1, diagnostic_U)

Integrated C* estimation (ejabT1E package)
```{r}
# Uses estimate_Cstar from ejabT1E: minimizes integrated squared deviation
# between empirical contradiction rate and theoretical T1E rate (closed-form
# objective), instead of pointwise (proportion - alpha/up)^2 per alpha.
# See: https://github.com/nathoogroup/ejabT1E
```

Prepare data for EJAB method
```{r}
rpcb_df$p = pmax(as.numeric(rpcb_df$Original.p.value), .Machine$double.eps)
rpcb_df$N = as.numeric(rpcb_df$Original.sample.size)
rpcb_df$q = as.numeric(rpcb_df$Original.df1)
is_chisq <- rpcb_df$Original.test.statistic.type == "chi-squared" & !is.na(rpcb_df$Original.test.statistic.type)
rpcb_df$q[is.na(rpcb_df$q) & is_chisq] <- as.numeric(rpcb_df$Original.df2[is.na(rpcb_df$q) & is_chisq])
rpcb_df$q[is.na(rpcb_df$q)] <- 1
rpcb_df <- rpcb_df[!is.na(rpcb_df$q), ]

rpcb_df <- rpcb_df[rpcb_df$N > 1 & rpcb_df$q >= 1, ]

cat("Number of rows:", nrow(rpcb_df), "\n")
cat("p range:", range(rpcb_df$p), "\n")
cat("N range:", range(rpcb_df$N), "\n")
cat("q range:", range(rpcb_df$q), "\n")
```

Run integrated C* estimation and produce the two-panel diagnostic figure
```{r calibration, fig.width=12, fig.height=5}
up <- 0.1  # Upper threshold u_p

# Compute eJAB for all studies (using ejabT1E)
rpcb_df$ejab <- ejab01(rpcb_df$p, rpcb_df$N, rpcb_df$q)

# Estimate C* via integrated objective (minimizes integrated squared deviation)
Cfit <- estimate_Cstar(rpcb_df$p, rpcb_df$ejab, up = up,
                      grid_range = c(0, 3), grid_n = 200)
Cstar <- Cfit$Cstar

# ── Panel (a): Calibration curve over alpha in (0, 0.05] ──
N_up <- sum(rpcb_df$p < up)
alpha_grid <- seq(0, alpha, length.out = 201)[-1]
emp_rate <- vapply(alpha_grid, function(a) {
  sum(rpcb_df$p < a & rpcb_df$ejab > Cstar) / N_up
}, numeric(1))
ref_rate <- alpha_grid / up

cal_df <- data.frame(alpha = alpha_grid, emp = emp_rate, ref = ref_rate)

p_cal <- ggplot(cal_df, aes(x = alpha)) +
  geom_line(aes(y = ref), color = "#D62728", linetype = "dashed", linewidth = 0.9) +
  geom_step(aes(y = emp), color = "#1F77B4", linewidth = 0.9) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.02))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = expression(alpha), y = "Empirical contradiction rate",
       title = "(a) Calibration curve") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(hjust = 0, face = "plain", size = 14),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "grey30"),
        axis.ticks = element_line(color = "grey30"))

# ── Panel (b): Q-Q diagnostic with simultaneous Beta band ──
idx_t1e <- which(rpcb_df$p < alpha & rpcb_df$ejab > Cstar)
U_vals <- diagnostic_U(rpcb_df$p[idx_t1e], rpcb_df$N[idx_t1e], rpcb_df$q[idx_t1e],
                       alpha, Cstar)

# Use ejabT1E's diagnostic_qqplot to compute band/intercept/slope, suppress its plot
qq_pdf <- tempfile(fileext = ".pdf"); pdf(qq_pdf); qq_obj <- diagnostic_qqplot(
  U_vals, alpha = alpha, Cstar = Cstar); dev.off(); unlink(qq_pdf)

qq_df <- data.frame(theoretical = qq_obj$theoretical,
                    observed    = qq_obj$observed,
                    lower       = qq_obj$lower,
                    upper       = qq_obj$upper)

p_qq <- ggplot(qq_df, aes(x = theoretical)) +
  geom_line(aes(y = lower), color = "grey50", linetype = "dashed", linewidth = 0.6) +
  geom_line(aes(y = upper), color = "grey50", linetype = "dashed", linewidth = 0.6) +
  geom_abline(intercept = qq_obj$intercept, slope = qq_obj$slope,
              color = "#D62728", linewidth = 0.9) +
  geom_point(aes(y = observed), color = "#1F77B4", size = 1.8) +
  labs(x = "Uniform(0, 1) quantiles",
       y = expression(paste("Sorted ", U[(i)])),
       title = "(b) Q-Q diagnostic") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(hjust = 0, face = "plain", size = 14),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "grey30"),
        axis.ticks = element_line(color = "grey30"))

diag_fig <- p_cal + p_qq + patchwork::plot_layout(widths = c(1, 1))

# Save and show
ggsave("rpcb_diagnostic.png", diag_fig, width = 12, height = 5, dpi = 300)
print(diag_fig)

cat("Integrated C* (alpha =", alpha, "):", round(Cstar, 4), "\n")
cat("Minimized objective:", round(Cfit$objective, 6), "\n")
cat(sprintf("QQ OLS: intercept = %.4f, slope = %.4f\n", qq_obj$intercept, qq_obj$slope))
cat(sprintf("Candidates outside 95%% simultaneous band: %d / %d\n",
            length(qq_obj$outside), length(U_vals)))
```

Compute f, U diagnostic, and flag Type I candidates using integrated C*
```{r}
# f_i: the p-value threshold at which eJAB = C* for each study's (n, q)
rpcb_df$f <- 1 - pchisq(
  (2 * rpcb_df$N^(1 / rpcb_df$q) / (rpcb_df$N^(1 / rpcb_df$q) - 1)) * log(sqrt(rpcb_df$N) / Cstar),
  df = rpcb_df$q
)

# Diagnostic U_i = (p_i - f_i) / (alpha - f_i)
rpcb_df$U_diagnostic <- diagnostic_U(rpcb_df$p, rpcb_df$N, rpcb_df$q, alpha, Cstar)

# Flag Type I candidates: p < alpha AND eJAB > C*
rpcb_df$type1_candidate <- (rpcb_df$p < alpha & rpcb_df$ejab > Cstar)
```

Summary and export
```{r}
# Summary
cat("\nType I candidates:\n")
print(rpcb_df[rpcb_df$type1_candidate, c("Paper..", "p", "ejab", "U_diagnostic", "replicated")])

# Compare replication rates
cat("\nReplication rate for Type I candidates:", 
    mean(rpcb_df$replicated[rpcb_df$type1_candidate]), "\n")
cat("Replication rate for non-Type I candidates:", 
    mean(rpcb_df$replicated[!rpcb_df$type1_candidate]), "\n")

# Export
write_csv(rpcb_df, "rpcb_output.csv")
```

Ground truth comparison (using replication as proxy)
```{r ground_truth}
# In RPP, ground truth is approximated by replication outcome:
#   "True T1E" = significant (p < alpha) AND failed to replicate
#   This is a proxy — some non-replicating studies may be underpowered rather
#   than true nulls, but it's the best available ground truth.

n_total     <- nrow(rpcb_df)
predicted   <- rpcb_df$type1_candidate
real_t1e    <- (rpcb_df$p < alpha) & (rpcb_df$replicated == 0)

n_real      <- sum(real_t1e)
n_predicted <- sum(predicted)
n_overlap   <- sum(real_t1e & predicted)

# Sensitivity: of real T1Es, how many did we catch?
sensitivity <- ifelse(n_real > 0, n_overlap / n_real, NA)

# Specificity: of non-T1Es, how many did we correctly leave alone?
n_not_t1e  <- n_total - n_real
n_true_neg <- n_not_t1e - (n_predicted - n_overlap)
specificity <- ifelse(n_not_t1e > 0, n_true_neg / n_not_t1e, NA)

# False positives: predicted T1E but actually replicated
n_false_pos <- n_predicted - n_overlap

# PPV (precision): of predicted T1Es, how many are real?
ppv <- ifelse(n_predicted > 0, n_overlap / n_predicted, NA)

cat("## Ground Truth Comparison\n")
cat(sprintf("- True T1Es (sig & did not replicate): %d / %d (%.1f%%)\n",
            n_real, n_total, 100 * n_real / n_total))
cat(sprintf("- Predicted T1Es: %d / %d (%.1f%%)\n",
            n_predicted, n_total, 100 * n_predicted / n_total))
cat(sprintf("- Correctly predicted: %d / %d (%.1f%%)\n",
            n_overlap, n_real,
            ifelse(n_real > 0, 100 * n_overlap / n_real, 0)))
cat(sprintf("- False positives: %d\n", n_false_pos))
cat(sprintf("- Sensitivity: %.3f\n", sensitivity))
cat(sprintf("- Specificity: %.3f\n", specificity))
cat(sprintf("- PPV (precision): %.3f\n", ppv))
```

Diagnostic QQ-Plot with OLS fit and simultaneous confidence band
```{r diagnostic_qqplot, fig.width=8, fig.height=7}
# ── Separate significant studies into detected Type I vs true positives ──
sig_df <- rpcb_df[rpcb_df$p < alpha, ]
n_sig  <- nrow(sig_df)

type1_df <- sig_df[sig_df$type1_candidate == TRUE, ]
tp_df    <- sig_df[sig_df$type1_candidate == FALSE, ]

n_type1    <- nrow(type1_df)
n_true_pos <- nrow(tp_df)

rep_type1    <- mean(type1_df$replicated)
rep_nontype1 <- mean(tp_df$replicated)

cat(sprintf("Significant studies: %d\n", n_sig))
cat(sprintf("  Detected Type I errors: %d  (replication rate %.1f%%)\n",
            n_type1, rep_type1 * 100))
cat(sprintf("  True positives:         %d  (replication rate %.1f%%)\n",
            n_true_pos, rep_nontype1 * 100))

# ── QQ-Plot using diagnostic_qqplot (OLS + MC-calibrated band) ──
type1_u   <- type1_df$U_diagnostic
type1_rep <- type1_df$replicated

# Save to PNG
png("rpp_diagnostic_qqplot.png", width = 8, height = 7, units = "in", res = 300)
qq_result <- diagnostic_qqplot(type1_u, alpha = alpha, Cstar = Cstar)
mtext(sprintf("%d detected Type I errors  |  replication rate %.0f%%",
              n_type1, rep_type1 * 100),
      side = 3, line = 0.3, cex = 0.85, col = "grey30")
dev.off()

# Also render in notebook
qq_result <- diagnostic_qqplot(type1_u, alpha = alpha, Cstar = Cstar)
mtext(sprintf("%d detected Type I errors  |  replication rate %.0f%%",
              n_type1, rep_type1 * 100),
      side = 3, line = 0.3, cex = 0.85, col = "grey30")

cat(sprintf("\nOLS fit: intercept = %.4f, slope = %.4f\n",
            qq_result$intercept, qq_result$slope))
if (length(qq_result$outside) > 0) {
  cat(sprintf("Points outside 95%% simultaneous band: %d\n",
              length(qq_result$outside)))
}
```

Overlaid distributions of U_i: Type I candidates vs true positives
```{r overlaid_distributions, fig.width=10, fig.height=5}
tp_u <- tp_df$U_diagnostic

# Clip true positives to reasonable range (remove extreme negatives)
tp_u_clipped <- tp_u[tp_u >= -1]

cat(sprintf("True positives clipped: keeping %d / %d with U_i >= -1\n",
            length(tp_u_clipped), length(tp_u)))

par(mfrow = c(1, 2))

# ── Overlaid histograms ──
brks <- seq(-1, 1, by = 0.1)

hist(tp_u_clipped, breaks = brks, freq = FALSE,
     col = adjustcolor("steelblue", 0.35), border = "steelblue",
     main = expression(paste("Overlaid ", U[i], " Distributions")),
     xlab = expression(U[i]), ylim = c(0, 3.5),
     xlim = c(-1, 1))
hist(type1_u, breaks = brks, freq = FALSE,
     col = adjustcolor("red", 0.4), border = "red",
     add = TRUE)
abline(h = 1, col = "black", lwd = 2, lty = 2)
legend("topleft",
       legend = c("Detected Type I", "True Positive", "Unif(0,1) density"),
       fill   = c(adjustcolor("red", 0.4), adjustcolor("steelblue", 0.35), NA),
       border = c("red", "steelblue", NA),
       lty    = c(NA, NA, 2), lwd = c(NA, NA, 2),
       bty = "n", cex = 0.85)

# ── Overlaid ECDFs ──
plot(ecdf(type1_u), pch = 19, cex = 0.5, col = "red",
     main = expression(paste("Overlaid ECDFs of ", U[i])),
     xlab = expression(U[i]), ylab = "Cumulative Probability",
     xlim = c(-1, 1))
lines(ecdf(tp_u_clipped), pch = 19, cex = 0.5, col = "steelblue")
abline(0, 1, col = "black", lwd = 2, lty = 2)
legend("topleft",
       legend = c("Detected Type I", "True Positive", "Unif(0,1) CDF"),
       col = c("red", "steelblue", "black"),
       lty = c(1, 1, 2), lwd = 2, bty = "n", cex = 0.85)

par(mfrow = c(1, 1))

# KS tests
cat("\nKS test — Type I candidates vs Unif(0,1):\n")
print(ks.test(type1_u, "punif"))
cat("\nKS test — True positives vs Unif(0,1):\n")
print(ks.test(tp_u, "punif"))
```

Pounds & Morris (2003) Beta-Uniform Mixture Model — exact implementation
```{r bum_library}
# Core BUM functions from Pounds & Morris (2003), adapted from the original
# bum-library.R released by Stan Pounds at St. Jude Children's Research Hospital.

dbum  <- function(x, a, lambda) lambda + (1 - lambda) * a * x^(a - 1)
pbum  <- function(x, a, lambda) lambda * x + (1 - lambda) * x^a

logit     <- function(x) log(x / (1 - x))
inv.logit <- function(x) exp(x) / (1 + exp(x))

neg.bum.logL <- function(pars, pvals) {
  a      <- inv.logit(pars[1])
  lambda <- inv.logit(pars[2])
  -sum(log(dbum(pvals, a, lambda)))
}

bum.mle <- function(pvals, nstartpts = 10) {
  adjeps   <- 0.5
  nadj     <- length(pvals)
  adjusted <- min(pvals) <= 0
  if (adjusted) pvals <- (nadj * pvals + adjeps) / (nadj + 2 * adjeps)

  starta      <- logit(runif(nstartpts + 1, 0.05, 0.95))
  startlambda <- logit(runif(nstartpts + 1, 0.05, 0.95))
  starta[1]      <- logit(0.5)
  startlambda[1] <- logit(0.5)

  bestlogL <- -Inf
  for (i in seq_along(starta)) {
    res <- tryCatch(
      nlm(neg.bum.logL, c(starta[i], startlambda[i]), pvals = pvals),
      error = function(e) NULL)
    if (!is.null(res) && -res$minimum > bestlogL) {
      a           <- inv.logit(res$estimate[1])
      lambda      <- inv.logit(res$estimate[2])
      logL        <- -res$minimum
      bestlogL    <- logL
      termination <- res$code
    }
  }
  list(a = a, lambda = lambda, logL = logL,
       pvals.adjusted = adjusted, termination = termination)
}

# Maximal uniform proportion extractable from BUM density (eq. 3 in paper)
ext.pi <- function(a, lambda) lambda + (1 - lambda) * a

# Empirical Bayes posterior P(alternative | p)
bum.emp.post <- function(x, a, lambda) {
  pi <- ext.pi(a, lambda)
  (dbum(x, a, lambda) - pi) / dbum(x, a, lambda)
}

# False discovery rate at threshold tau
bum.FDR <- function(tau, a, lambda) {
  pi <- ext.pi(a, lambda)
  tau * pi / pbum(tau, a, lambda)
}

# Proportions of error types at threshold tau
bum.false.positive <- function(tau, a, lambda) tau * ext.pi(a, lambda)
bum.true.positive  <- function(tau, a, lambda) pbum(tau, a, lambda) - tau * ext.pi(a, lambda)
bum.false.negative <- function(tau, a, lambda) 1 - pbum(tau, a, lambda) - ext.pi(a, lambda) * (1 - tau)
bum.true.negative  <- function(tau, a, lambda) ext.pi(a, lambda) * (1 - tau)

# p-value threshold for a desired FDR level
find.FDR.threshold <- function(fdr, a, lambda) {
  pi <- ext.pi(a, lambda)
  ((pi - fdr * lambda) / (fdr * (1 - lambda)))^(1 / (a - 1))
}

# p-value threshold for a desired EB posterior level
find.EB.threshold <- function(EB, a, lambda) {
  pi <- ext.pi(a, lambda)
  ((EB * lambda + a * (1 - lambda)) / (a * (1 - EB) * (1 - lambda)))^(1 / (a - 1))
}
```

Fit BUM via MLE to RPP p-values
```{r bum_fit}
set.seed(42)
bum_fit <- bum.mle(rpcb_df$p, nstartpts = 20)

pi_hat <- ext.pi(bum_fit$a, bum_fit$lambda)

cat("Pounds & Morris BUM — MLE Estimates:\n")
cat(sprintf("  a (Beta shape):       %.4f\n", bum_fit$a))
cat(sprintf("  lambda (uniform wt):  %.4f\n", bum_fit$lambda))
cat(sprintf("  pi (extracted null):  %.4f  [= lambda + (1-lambda)*a]\n", pi_hat))
cat(sprintf("  log-likelihood:       %.2f\n", bum_fit$logL))
cat(sprintf("  convergence code:     %d\n", bum_fit$termination))

fdr_at_alpha <- bum.FDR(alpha, bum_fit$a, bum_fit$lambda)
cat(sprintf("\nAt alpha = %.2f:\n", alpha))
cat(sprintf("  FDR(alpha):                  %.4f\n", fdr_at_alpha))
cat(sprintf("  Est. false positive rate:    %.4f\n",
            bum.false.positive(alpha, bum_fit$a, bum_fit$lambda)))
cat(sprintf("  Est. true positive rate:     %.4f\n",
            bum.true.positive(alpha, bum_fit$a, bum_fit$lambda)))
```

BUM model fit visualization
```{r bum_fit_plot, fig.width=10, fig.height=5}
par(mfrow = c(1, 2))

# Histogram with BUM density overlay
h <- hist(rpcb_df$p, breaks = 30, plot = FALSE)
hist(rpcb_df$p, breaks = 30, freq = FALSE,
     col = adjustcolor("grey80", 0.5), border = "grey50",
     main = "Pounds & Morris BUM Fit",
     xlab = "p-value",
     ylim = c(0, max(h$density) * 1.3))
curve(dbum(x, bum_fit$a, bum_fit$lambda),
      from = 0.001, to = 1, add = TRUE, col = "red", lwd = 2, n = 500)
abline(h = pi_hat, col = "blue", lty = 2, lwd = 1.5)
legend("topright",
       legend = c(
         sprintf("BUM density (a=%.3f, lam=%.3f)", bum_fit$a, bum_fit$lambda),
         sprintf("Extracted null pi=%.3f", pi_hat)),
       col = c("red", "blue"), lty = c(1, 2), lwd = c(2, 1.5), bty = "n", cex = 0.8)

# Error region diagram
n_total_studies <- nrow(rpcb_df)
fp_prop <- bum.false.positive(alpha, bum_fit$a, bum_fit$lambda)
tp_prop <- bum.true.positive(alpha, bum_fit$a, bum_fit$lambda)
fn_prop <- bum.false.negative(alpha, bum_fit$a, bum_fit$lambda)
tn_prop <- bum.true.negative(alpha, bum_fit$a, bum_fit$lambda)

barplot(c(FP = fp_prop, TP = tp_prop, FN = fn_prop, TN = tn_prop) * n_total_studies,
        col = c("red", "green3", "orange", "steelblue"),
        main = sprintf("BUM Error Partition at alpha=%.2f", alpha),
        ylab = "Estimated count", las = 1)
legend("topright",
       legend = sprintf(c("FP: %.1f", "TP: %.1f", "FN: %.1f", "TN: %.1f"),
                        c(fp_prop, tp_prop, fn_prop, tn_prop) * n_total_studies),
       fill = c("red", "green3", "orange", "steelblue"), bty = "n", cex = 0.8)

par(mfrow = c(1, 1))
```

BUM-based Type I error detection using EB posterior (exact P&M approach)
```{r bum_classification}
# Pounds & Morris EB posterior: P(alternative | p_i)
# A study is classified as null (Type I error) when EB posterior < 0.5,
# meaning it is more likely null than alternative.
rpcb_df$eb_post <- bum.emp.post(rpcb_df$p, bum_fit$a, bum_fit$lambda)

# EB threshold: the p-value below which EB posterior > 0.5
eb_threshold <- find.EB.threshold(0.5, bum_fit$a, bum_fit$lambda)

cat(sprintf("EB posterior threshold (P(alt|p) = 0.5) at p = %.6f\n", eb_threshold))
cat(sprintf("  alpha = %.4f\n", alpha))

if (eb_threshold < alpha) {
  cat(sprintf("  eb_threshold < alpha: significant studies with p in (%.6f, %.4f]\n",
              eb_threshold, alpha))
  cat("  are classified as more likely null than alternative.\n")
} else {
  cat("  eb_threshold >= alpha: all significant studies have EB posterior > 0.5,\n")
  cat("  so the BUM classifies all of them as likely alternatives.\n")
}

# Predicted T1Es: significant AND EB posterior < 0.5
rpcb_df$pm_type1 <- (rpcb_df$p < alpha) & (rpcb_df$eb_post < 0.5)

# Also flag using FDR-based count: expected # of FPs among significant studies
n_sig <- sum(rpcb_df$p < alpha)
n_expected_fp <- round(fdr_at_alpha * n_sig)

cat(sprintf("\n--- Classification Results ---\n"))
cat(sprintf("Significant studies:              %d\n", n_sig))
cat(sprintf("BUM EB-posterior T1E (post<0.5):  %d\n", sum(rpcb_df$pm_type1)))
cat(sprintf("BUM FDR-expected T1Es:            %d  (FDR=%.3f x %d)\n",
            n_expected_fp, fdr_at_alpha, n_sig))
cat(sprintf("eJAB predicted T1Es:              %d\n", sum(rpcb_df$type1_candidate)))

# EB posterior summary for significant studies
sig_eb <- rpcb_df$eb_post[rpcb_df$p < alpha]
cat(sprintf("\nEB posterior summary (significant studies):\n"))
print(summary(sig_eb))
```

Head-to-head comparison: eJAB vs Pounds & Morris BUM
```{r head_to_head}
# --- Direct EB posterior classification (the exact P&M approach) ---
pm_predicted   <- rpcb_df$pm_type1
pm_n_predicted <- sum(pm_predicted)
pm_n_overlap   <- sum(real_t1e & pm_predicted)
pm_n_false_pos <- pm_n_predicted - pm_n_overlap

pm_sensitivity <- ifelse(n_real > 0, pm_n_overlap / n_real, NA)
pm_n_true_neg  <- (n_total - n_real) - pm_n_false_pos
pm_specificity <- ifelse(n_not_t1e > 0, pm_n_true_neg / n_not_t1e, NA)
pm_ppv         <- ifelse(pm_n_predicted > 0, pm_n_overlap / pm_n_predicted, NA)

# --- FDR-budget comparison: flag top-k by lowest EB posterior, k = FDR-expected count ---
sig_rows <- which(rpcb_df$p < alpha)
rpcb_df$pm_fdr_flag <- FALSE

if (n_expected_fp > 0 && length(sig_rows) >= n_expected_fp) {
  eb_sig      <- rpcb_df$eb_post[sig_rows]
  top_k       <- order(eb_sig)[1:n_expected_fp]
  fdr_flag_idx <- sig_rows[top_k]
  rpcb_df$pm_fdr_flag[fdr_flag_idx] <- TRUE
}

fdr_n_predicted <- sum(rpcb_df$pm_fdr_flag)
fdr_n_overlap   <- sum(real_t1e & rpcb_df$pm_fdr_flag)
fdr_n_fp        <- fdr_n_predicted - fdr_n_overlap
fdr_sens        <- ifelse(n_real > 0, fdr_n_overlap / n_real, NA)
fdr_spec        <- ifelse(n_not_t1e > 0, (n_not_t1e - fdr_n_fp) / n_not_t1e, NA)
fdr_ppv         <- ifelse(fdr_n_predicted > 0, fdr_n_overlap / fdr_n_predicted, NA)
fdr_rep         <- ifelse(fdr_n_predicted > 0, mean(rpcb_df$replicated[rpcb_df$pm_fdr_flag]), NA)

comparison_df <- data.frame(
  Metric = c("Predicted T1Es", "Correctly Identified", "False Positives",
             "Sensitivity", "Specificity", "PPV (Precision)", "Replication Rate"),
  eJAB = c(n_predicted, n_overlap, n_false_pos,
           round(sensitivity, 3), round(specificity, 3), round(ppv, 3),
           round(mean(rpcb_df$replicated[rpcb_df$type1_candidate]), 3)),
  BUM_EB = c(pm_n_predicted, pm_n_overlap, pm_n_false_pos,
             round(pm_sensitivity, 3), round(pm_specificity, 3), round(pm_ppv, 3),
             ifelse(pm_n_predicted > 0,
                    round(mean(rpcb_df$replicated[rpcb_df$pm_type1]), 3), NA)),
  BUM_FDR_budget = c(fdr_n_predicted, fdr_n_overlap, fdr_n_fp,
                     round(fdr_sens, 3), round(fdr_spec, 3), round(fdr_ppv, 3),
                     round(fdr_rep, 3))
)

cat("\n## Head-to-Head: eJAB vs Pounds & Morris BUM\n")
cat("   BUM_EB:         EB posterior < 0.5 (exact P&M classification)\n")
cat("   BUM_FDR_budget: top-k by lowest EB posterior, k = FDR*n_sig\n\n")
print(comparison_df, row.names = FALSE)

# McNemar's test comparing eJAB vs BUM FDR-budget classification
b_fdr <- sum(rpcb_df$type1_candidate & !rpcb_df$pm_fdr_flag)
c_fdr <- sum(!rpcb_df$type1_candidate & rpcb_df$pm_fdr_flag)
cat(sprintf("\nMcNemar's test — eJAB vs BUM_FDR_budget (eJAB-only=%d, BUM-only=%d):\n",
            b_fdr, c_fdr))
if (b_fdr + c_fdr > 0) {
  print(mcnemar.test(table(rpcb_df$type1_candidate, rpcb_df$pm_fdr_flag)))
} else {
  cat("  Methods agree on all studies.\n")
}
```

ROC comparison among significant studies
```{r roc_comparison, fig.width=8, fig.height=7}
sig_idx   <- which(rpcb_df$p < alpha)
sig_truth <- real_t1e[sig_idx]

# eJAB: higher = more likely null
score_ejab <- rpcb_df$ejab[sig_idx]
# BUM EB posterior: LOWER = more likely null, so negate for consistent direction
score_bum  <- -rpcb_df$eb_post[sig_idx]

simple_roc <- function(score, truth) {
  ord   <- order(score, decreasing = TRUE)
  score <- score[ord]
  truth <- truth[ord]

  tp    <- cumsum(truth)
  fp    <- cumsum(!truth)
  n_pos <- sum(truth)
  n_neg <- sum(!truth)

  tpr <- c(0, tp / n_pos, 1)
  fpr <- c(0, fp / n_neg, 1)
  data.frame(fpr = fpr, tpr = tpr)
}

trap_auc <- function(roc) {
  n <- nrow(roc)
  sum(diff(roc$fpr) * (roc$tpr[-1] + roc$tpr[-n]) / 2)
}

roc_ejab <- simple_roc(score_ejab, sig_truth)
roc_bum  <- simple_roc(score_bum,  sig_truth)
auc_ejab <- trap_auc(roc_ejab)
auc_bum  <- trap_auc(roc_bum)

plot(roc_ejab$fpr, roc_ejab$tpr, type = "l", col = "red", lwd = 2,
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     main = "ROC: eJAB vs Pounds & Morris BUM\n(among significant studies)")
lines(roc_bum$fpr, roc_bum$tpr, col = "blue", lwd = 2)
abline(0, 1, lty = 2, col = "grey50")
legend("bottomright",
       legend = c(sprintf("eJAB (AUC = %.3f)", auc_ejab),
                  sprintf("BUM  (AUC = %.3f)", auc_bum)),
       col = c("red", "blue"), lwd = 2, bty = "n")

cat(sprintf("AUC — eJAB: %.4f\n", auc_ejab))
cat(sprintf("AUC — BUM:  %.4f\n", auc_bum))
```

eJAB vs BUM EB posterior scatter plot for significant studies
```{r scatter_comparison, fig.width=9, fig.height=6}
sig_df_ext <- rpcb_df[rpcb_df$p < alpha, ]

plot(sig_df_ext$ejab, sig_df_ext$eb_post,
     pch = 16, cex = 0.9,
     col = ifelse(sig_df_ext$replicated == 1,
                  adjustcolor("steelblue", 0.6),
                  adjustcolor("red", 0.6)),
     xlab = "eJAB statistic",
     ylab = "BUM EB posterior P(alt | p)",
     main = "eJAB vs P&M EB Posterior (significant studies)")

abline(h = 0.5, lty = 2, col = "grey50")
abline(v = Cstar, lty = 2, col = "grey50")

text(x = max(sig_df_ext$ejab) * 0.95, y = 0.53,
     "EB post = 0.5", cex = 0.65, col = "grey40", pos = 2)
text(x = Cstar, y = max(sig_df_ext$eb_post) * 0.98,
     sprintf("C* = %.3f", Cstar), cex = 0.65, col = "grey40", pos = 4)

legend("bottomright",
       legend = c("Replicated", "Did not replicate"),
       col = c("steelblue", "red"), pch = 16, bty = "n")

# Quadrant analysis: eJAB>C* flags T1E; EB post < 0.5 flags T1E
q_both    <- sum(sig_df_ext$ejab > Cstar & sig_df_ext$eb_post < 0.5)
q_ejab    <- sum(sig_df_ext$ejab > Cstar & sig_df_ext$eb_post >= 0.5)
q_bum     <- sum(sig_df_ext$ejab <= Cstar & sig_df_ext$eb_post < 0.5)
q_neither <- sum(sig_df_ext$ejab <= Cstar & sig_df_ext$eb_post >= 0.5)

cat(sprintf("\nQuadrant Analysis (significant studies, n=%d):\n", nrow(sig_df_ext)))
cat(sprintf("  Both flag T1E  (eJAB>C* & EB<0.5):  %d\n", q_both))
cat(sprintf("  eJAB only      (eJAB>C* & EB>=0.5): %d\n", q_ejab))
cat(sprintf("  BUM only       (eJAB<=C* & EB<0.5): %d\n", q_bum))
cat(sprintf("  Neither flags  (eJAB<=C* & EB>=0.5):%d\n", q_neither))

if (q_both > 0)
  cat(sprintf("  Rep rate — both flag:    %.1f%%\n",
              mean(sig_df_ext$replicated[sig_df_ext$ejab > Cstar &
                                         sig_df_ext$eb_post < 0.5]) * 100))
if (q_ejab > 0)
  cat(sprintf("  Rep rate — eJAB only:    %.1f%%\n",
              mean(sig_df_ext$replicated[sig_df_ext$ejab > Cstar &
                                         sig_df_ext$eb_post >= 0.5]) * 100))
if (q_bum > 0)
  cat(sprintf("  Rep rate — BUM only:     %.1f%%\n",
              mean(sig_df_ext$replicated[sig_df_ext$ejab <= Cstar &
                                         sig_df_ext$eb_post < 0.5]) * 100))
if (q_neither > 0)
  cat(sprintf("  Rep rate — neither:      %.1f%%\n",
              mean(sig_df_ext$replicated[sig_df_ext$ejab <= Cstar &
                                         sig_df_ext$eb_post >= 0.5]) * 100))
```
