---
title: "Simulation: eJAB vs AIC vs BIC for Model Selection"
output: html_document
date: "2025-02-18"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)
```

## Goal

Compare **eJAB**, **AIC**, and **BIC** in a model-selection simulation across two
scenarios: (A) selecting the correct model when the true model has two
predictors, and (B) avoiding overfitting when extra noise predictors are
available.

## eJAB function (from RPP)

```{r ejab-def}
ejab01 <- function(p, n, q) {
  stopifnot(all(p > 0 & p < 1), all(n > 1), all(q >= 1))
  chi_quant <- qchisq(1 - p, df = q)
  sqrt(n) * exp(-0.5 * (n^(1/q) - 1) / n^(1/q) * chi_quant)
}
```

**Interpretation refresher**: eJAB is a Bayes-factor-like quantity measuring
evidence *for the null*.

- **Small eJAB** (< 1) = strong evidence the alternative (larger model) is real.
- **Large eJAB** (> 1) = the null / simpler model is favored.

In the RPP pipeline, `ejab > C*` flags suspected Type I errors — studies where
the p-value looks significant but the Bayes factor actually favors the null.

## Helpers

```{r helpers}
clamp_p <- function(p) pmax(pmin(p, 1 - 1e-15), 1e-15)

generate_data <- function(n, betas, sigma = 1, n_noise = 0) {
  k <- length(betas) - 1
  X <- matrix(rnorm(n * (k + n_noise)), n, k + n_noise)
  colnames(X) <- paste0("x", seq_len(k + n_noise))
  y <- X[, seq_len(k), drop = FALSE] %*% betas[-1] + betas[1] + rnorm(n, 0, sigma)
  data.frame(y = as.numeric(y), X)
}
```

## Scenario A — True model with 2 predictors

**True model**: $y = 0.3 x_1 + 0.3 x_2 + \varepsilon$.
Candidates: null, $x_1$, $x_2$, $x_1 + x_2$.

```{r scenario-a-functions}
fit_A <- function(d) {
  list(
    null = lm(y ~ 1,      data = d),
    M1   = lm(y ~ x1,     data = d),
    M2   = lm(y ~ x2,     data = d),
    M12  = lm(y ~ x1+x2,  data = d)
  )
}

select_aic <- function(fits) which.min(sapply(fits, AIC))
select_bic <- function(fits) which.min(sapply(fits, BIC))

# eJAB: compare each non-null model to null; select the one with the smallest
# eJAB (strongest evidence against null).  If all eJAB >= 1, select null.
select_ejab <- function(fits, n) {
  null_fit <- fits[["null"]]
  non_null <- fits[names(fits) != "null"]
  qs <- sapply(non_null, function(f) length(coef(f)) - length(coef(null_fit)))

  pvals <- sapply(non_null, function(f) {
    clamp_p(anova(null_fit, f)$`Pr(>F)`[2])
  })

  ejabs <- mapply(ejab01, p = pvals, n = n, q = qs)
  if (all(ejabs >= 1)) return(1L)
  1L + which.min(ejabs)
}

model_names_A <- c("null", "M1", "M2", "M12")
true_idx_A    <- 4L
```

```{r run-a}
n_sims  <- 500
n_vec   <- c(25, 50, 100, 200)
res_A   <- list()

for (nn in n_vec) {
  counts <- matrix(0L, nrow = 4, ncol = 3,
                   dimnames = list(model_names_A, c("AIC", "BIC", "eJAB")))
  for (i in seq_len(n_sims)) {
    d    <- generate_data(nn, betas = c(0, 0.3, 0.3))
    fits <- fit_A(d)
    counts[select_aic(fits),         "AIC"]  <- counts[select_aic(fits),  "AIC"]  + 1L
    counts[select_bic(fits),         "BIC"]  <- counts[select_bic(fits),  "BIC"]  + 1L
    counts[select_ejab(fits, nn),    "eJAB"] <- counts[select_ejab(fits, nn), "eJAB"] + 1L
  }
  res_A[[as.character(nn)]] <- counts
}
```

### Correct-selection rate (true model = M12)

```{r table-a}
tab_A <- do.call(rbind, lapply(names(res_A), function(nn) {
  m <- res_A[[nn]]
  data.frame(
    n    = as.integer(nn),
    AIC  = sprintf("%.1f%%", 100 * m["M12","AIC"]  / n_sims),
    BIC  = sprintf("%.1f%%", 100 * m["M12","BIC"]  / n_sims),
    eJAB = sprintf("%.1f%%", 100 * m["M12","eJAB"] / n_sims)
  )
}))
knitr::kable(tab_A, align = "r",
             caption = "Scenario A — % selecting true model (M12) out of 500 sims")
```

### Full choice distributions

```{r choices-a}
for (nn in names(res_A)) {
  cat("\n### n =", nn, "\n")
  print(res_A[[nn]])
}
```

## Scenario B — Overfitting test (true model = null)

**True model**: $y = \varepsilon$ (no signal). Five candidate predictors are all
noise. This tests whether each criterion correctly favors the null.

```{r scenario-b}
fit_B <- function(d) {
  list(
    null = lm(y ~ 1, data = d),
    M1   = lm(y ~ x1, data = d),
    M2   = lm(y ~ x1+x2, data = d),
    M3   = lm(y ~ x1+x2+x3, data = d),
    M4   = lm(y ~ x1+x2+x3+x4, data = d),
    M5   = lm(y ~ x1+x2+x3+x4+x5, data = d)
  )
}

select_ejab_B <- function(fits, n) {
  null_fit <- fits[["null"]]
  non_null <- fits[names(fits) != "null"]
  qs <- sapply(non_null, function(f) length(coef(f)) - length(coef(null_fit)))
  pvals <- sapply(non_null, function(f) clamp_p(anova(null_fit, f)$`Pr(>F)`[2]))
  ejabs <- mapply(ejab01, p = pvals, n = n, q = qs)
  if (all(ejabs >= 1)) return(1L)
  1L + which.min(ejabs)
}

model_names_B <- c("null", paste0("M", 1:5))

res_B <- list()
for (nn in n_vec) {
  aic_c <- bic_c <- ejab_c <- setNames(integer(6), model_names_B)
  for (i in seq_len(n_sims)) {
    d    <- generate_data(nn, betas = c(0, rep(0, 5)), n_noise = 0)
    fits <- fit_B(d)
    a <- select_aic(fits);           aic_c[a]  <- aic_c[a]  + 1L
    b <- select_bic(fits);           bic_c[b]  <- bic_c[b]  + 1L
    e <- select_ejab_B(fits, nn);    ejab_c[e] <- ejab_c[e] + 1L
  }
  res_B[[as.character(nn)]] <- list(AIC = aic_c, BIC = bic_c, eJAB = ejab_c)
}
```

### Null-selection rate (true model = null, no signal)

```{r table-b}
tab_B <- do.call(rbind, lapply(names(res_B), function(nn) {
  r <- res_B[[nn]]
  data.frame(
    n     = as.integer(nn),
    AIC   = sprintf("%.1f%%", 100 * r$AIC["null"] / n_sims),
    BIC   = sprintf("%.1f%%", 100 * r$BIC["null"] / n_sims),
    eJAB  = sprintf("%.1f%%", 100 * r$eJAB["null"] / n_sims)
  )
}))
knitr::kable(tab_B, align = "r",
             caption = "Scenario B — % correctly selecting null (no signal) out of 500 sims")
```

### Full choice distributions

```{r choices-b}
for (nn in names(res_B)) {
  r <- res_B[[nn]]
  cat("\n### n =", nn, "\n")
  cat("AIC:  ", paste(model_names_B, r$AIC,  sep="=", collapse=", "), "\n")
  cat("BIC:  ", paste(model_names_B, r$BIC,  sep="=", collapse=", "), "\n")
  cat("eJAB: ", paste(model_names_B, r$eJAB, sep="=", collapse=", "), "\n")
}
```

## Summary

```{r summary-plot, fig.width = 9, fig.height = 4}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Panel A — correct selection (true model has signal)
pct_A <- sapply(res_A, function(m) {
  c(AIC = m["M12","AIC"], BIC = m["M12","BIC"],
    eJAB = m["M12","eJAB"]) / n_sims * 100
})
colnames(pct_A) <- n_vec
barplot(pct_A, beside = TRUE, ylim = c(0, 100),
        col = c("steelblue","coral","seagreen"),
        xlab = "n", ylab = "% correct",
        main = "A: True model selection (M12)")
legend("topleft", rownames(pct_A), fill = c("steelblue","coral","seagreen"),
       cex = 0.75, bty = "n")

# Panel B — null retention (no signal)
pct_B <- sapply(res_B, function(r) {
  c(AIC = r$AIC["null"], BIC = r$BIC["null"], eJAB = r$eJAB["null"]) / n_sims * 100
})
colnames(pct_B) <- n_vec
barplot(pct_B, beside = TRUE, ylim = c(0, 100),
        col = c("steelblue","coral","seagreen"),
        xlab = "n", ylab = "% selecting null",
        main = "B: Null retention (no signal)")
legend("topleft", rownames(pct_B), fill = c("steelblue","coral","seagreen"),
       cex = 0.75, bty = "n")
```

### Take-aways

- **AIC** is aggressive — good at finding the true model when signal exists, but
  prone to overfitting noise predictors (Scenario B).
- **BIC** adds a stronger penalty ($\log n$) and resists overfitting better, at
  the cost of lower power in small samples.
- **eJAB** acts as a Bayes-factor filter: it selects the model only when evidence
  is decisive (eJAB < 1 against the null). This makes it conservative — it may
  under-select in small $n$, but it strongly resists overfitting, similar to or
  better than BIC.

Adjust `betas` and `sigma` to explore different signal-to-noise regimes.
